---
title: "Midterm"
author: Justin Gould
format: pdf
editor: visual
---

**Installed Packages:**

To begin, I loaded the tidyverse package and set the working directory to where my data was located. Lastly, I created a dataframe called housing_data to store the data. Next, I performed an exploratory data analysis on the housing data with the goal of being able to fit this data into regression models and predict housing prices using the variable SalePrice.

```{r}
#| warning: false 
#| message: false
#| echo: true
library(tidyverse)
setwd("C:/Users/justi/OneDrive/Desktop/Grad School/UTSA 1st semester/STA 6443 Statistical Modeling/Midterm") 

housing_data <- read.csv("ames_housing_data-1.csv")  
```

## 

## 1. Univariate Analysis

To begin my EDA I created a data frame called eda with the purpose of narrowing down the variables within the housing dataset. I restricted my analysis to only focus on these variables: SalePrice, GrLivArea, LotArea, OverallQual, YearBuilt, GarageCars, Neighborhood, CentralAir. I renamed this revised dataset "eda."

```{r}
#| warning: false 
#| message: false
#| echo: true
eda <- housing_data |> select(SalePrice, GrLivArea, LotArea, OverallQual, YearBuilt, GarageCars, Neighborhood, CentralAir)

```

## 

### Data Dictionary

**SalePrice:** The price the house sold for (dependent variable)

**GrLivArea:** Above grade (ground) living area square feet

**LotArea:** Lot size in square feet

**OverallQual:** Rates the overall material and finish of the house (1-10 scale)

**YearBuilt:** Original construction date

**GarageCars:** Size of garage in car capacity

**Neighborhood:** Physical locations within Ames city limits (categorical)

**CentralAir:** Central air conditioning (Yes or No)

## 

Next, I ran a histogram plot to observe the distribution of the our dependant variable in this analysis. **Sale price will be the dependant variable for this analysis**

```{r}
#| warning: false 
#| message: false
#| echo: true
eda |> ggplot(aes(x=SalePrice)) + 
        geom_histogram(fill = "darkgreen", color = "black") + 
        labs(title = "Sales Price Distribution")
```

**Takeaway:** This graph uncovers that this data is *right skewed*, meaning that the majority of sales price values are in the low to moderate range within this dataset

## 

Next, I evaluated the distribution of the independent variables.

```{r}
#| warning: false 
#| message: false
#| echo: true
eda |> ggplot(aes(x=OverallQual)) + 
        geom_histogram(fill = "darkgreen", color = "black", bins = 30) + 
        labs(
          title = "Overall Quality Distribution",
          x = "Overall Quality",
          y = "Count"
          )
```

**Takeaway:** This graph shows that we will be working with a majority of houses in the 5-7 point quality range.

## 

```{r}
#| warning: false 
#| message: false
#| echo: true
 eda |> ggplot(aes(x=YearBuilt)) + 
        geom_histogram(fill = "darkgreen", color = "black") + 
        labs(
    title = "Year Built Distribution",
    x = "Year Built",
    y = "Count"
  ) +
  scale_x_continuous(breaks = seq(1870, 2020, by = 20))
```

**Takeaway:** This graph shows a pretty even number of houses built in the 50s-70s and then uncovers a massive spike in houses built in the early 2000s.

## 

```{r}
#| warning: false 
#| message: false
#| echo: true
eda |> 
    filter(LotArea <= 50000) |> 
    ggplot(aes(x=LotArea)) + 
        geom_histogram(fill = "darkgreen", color = "black", bins = 30) + 
        labs(
      title = "Lot Area Distribution", 
      x = "Lot Area",
      y = "Count"
  ) + scale_x_continuous(
    breaks = seq(0, max(eda$LotArea, na.rm = TRUE), by = 10000))
```

**Takeaway:** The lot area data is normally distributed in one section and then it right skews. There are many outliers in the 40,000- 100,000 area range but I have applied the filter to normalize the data.

## 

Next, I ran a summary function to evaluate the dependant variable and the categorical variables further.

```{r}
#| warning: false 
#| message: false
#| echo: true
eda |> summary(SalePrice)
```

**Takeaway:** The SalePrice mean is 214,000 with the min being 34k and the max being 755k. This wide range shows that this sales price data is going to show us a wide variation of sales prices throughout our analysis.

## 

## 2. Bivariet Analysis

In phase 2 of my EDA, I investigate the relationships between the dependent variable and the independent variables.

```{r}
#| warning: false 
#| message: false
#| echo: true
cor(eda$SalePrice, eda$GrLivArea)
cor(eda$SalePrice, eda$OverallQual)
cor(eda$SalePrice, eda$LotArea)
cor(eda$SalePrice, eda$GarageCars)
cor(eda$SalePrice, eda$YearBuilt)
cor(eda$OverallQual, eda$GarageCars)
```

**Takeaway:** The strongest correlation is between Sale Price and Overall Quality of **.79**. I also discovered a medium correlation between two independent variables Overall Quality and GarageCars of .64.

## 

I explore our strongest correlation between Sale Price and Overall quality here...

```{r}
#| warning: false 
#| message: false
#| echo: true
eda |> ggplot(aes(x=OverallQual, y=SalePrice)) + 
    geom_point(color="darkgreen") +
    geom_smooth(method = "lm", se = FALSE, color = "black") +
    labs(title = "Relationship between Sale Price and Overall Quality")

```

**Takeaway:** As quality score increase on the 1-10 scale, sale price increases in a linear fashion with it.

## 

I explore our next strongest corellation here by filtering out the GrLivArea to less than 3,000 sqft to eliminate outliers.

```{r}
#| warning: false 
#| message: false
#| echo: true
eda |> filter(GrLivArea <= 3000) |> 
    ggplot(aes(x=SalePrice, y=GrLivArea)) + 
    geom_point(color = "darkblue") +
    labs(
            title = "Relationship between Sale Price and Above grade (ground) living area"
          )
```

**Takeaway:** This data reveals that Sale Price and Above grade (ground) living area are positively correlated with most of the datapoints moving in linear fasion.

## 

Next, I evaluate the relationship between our dependant variable and one of the categorical variables, cenralair.

```{r}
#| warning: false 
#| message: false
#| echo: true
eda |> ggplot(aes(x=CentralAir, y=SalePrice)) + 
  geom_boxplot(fill = "lightgreen", color = "black") +
    labs(title = "Relationship between Sale Price and Air Conditioning")

```

**Takeaway:** Most of the high value properties contrain central air conditioning units.

## 

Next, I investigate the relationship between Sale Price, GrLivArea, and Overall Quality.

```{r}
#| warning: false 
#| message: false
#| echo: true
 eda |> ggplot(aes(x = GrLivArea, y = SalePrice, color= OverallQual)) +
    geom_point() +
    labs(title = "Sale price relationship between GrLivArea and Overall Quality")
```

**Takeaway:** This graph shows that the larger the above ground living area is, the better the quality, AND the higher the sales price tend to be.

## 

## 3. Data Cleaning and Transformation

For data cleaning, I checked for outliers within the categorical variables, null values, and references the above charts to take note of any outliers. The outliers I found were in GrLivArea and Lot Area.

```{r}
#| warning: false 
#| message: false
#| echo: true
 
eda |> summarize(CentralAir_nulls = sum(is.na(CentralAir)), Neighborhood_nulls = sum(is.na(Neighborhood)))
colSums(is.na(eda))

```

Additionally, I changed row names in the neighborhood column for uniformity.

```{r}
#| warning: false 
#| message: false
#| echo: true
 unique(eda$Neighborhood)
eda <- eda |> mutate(Neighborhood = str_to_title(Neighborhood))
```

## 

## 4. Feature Selection

I believe based off their high correlation with Sale price, that GrLivArea and OverallQual will be the most important variables for predicting sale price in a model. The positive linear relationship that these variables showed in the bivariate analysis prove that these variables can be accurate predictors for sale price.

## 

## Part 2.

I added a column to the EDA dataset to conduct a one way ANOVA test called Location. This location column shows the neighborhoods categorized as North, South, or West.

```{r}
#| warning: false 
#| message: false
#| echo: true
 
eda2.0 <- eda |> group_by(Neighborhood) |> 
       mutate(Location = case_when(
        Neighborhood %in% c(
                        "Blmngtn", "Brdale", "Brkside", "Gilbert", "Idotrr",
                        "Names", "Noridge", "Npkvill", "Nridght",
                        "Nwames", "Oldtown", "Somerst", "Stonebr") ~ "North",
        Neighborhood %in% c(
                        "Meadowv", "Mitchel", "Timber") ~ "South",
        Neighborhood %in% c(
                        "Blueste", "Clearcr", "Collgcr", "Crawfor", "Edwards",
                        "Swisu", "Sawyer", "Sawyerw", "Veenker") ~ "West"))
```

## 

## AVOVA

**Purpose:** The purpose of this ANOVA test will be to determine if there is a statistically significant difference in sales between the North, South, and West locations.

**Null Hypotheses:**

*There is no difference in mean Sale Price across all locations (North, South, West).*

## 

**Alternative Hypotheses:**

*There is a difference in mean across at least one location.*

First off, I used as.factor to change the categorical variables to factors. Next, I gathered the means of saleprice across location.

```{r}
#| warning: false 
#| message: false
#| echo: true 
eda2.0$Neighborhood<-as.factor(eda2.0$Neighborhood)
eda2.0 |> group_by(Location) |> summarize(mean_location = mean(SalePrice))

```

Next, I created a boxplot to show North, South, and West group means.

```{r}
#| warning: false 
#| message: false
#| echo: true
eda2.0 |> ggplot(aes(x = Location, y = SalePrice)) +
          geom_boxplot(fill = "lightblue", color = "black") +
          labs(title = "Location Group Means",
               x = "Sale Price",
               y = "Location") +
          theme_minimal()
```

## 

Next, I conducted the ANOVA test and a Tukeys test to be sure of the results.

```{r}
#| warning: false 
#| message: false
#| echo: true
oneway <- aov(SalePrice ~ Location , data = eda2.0)
summary(oneway)
```

## 

**Conclusion:** We can conclude from the P \> .0271 value that the P value is significant As a result, to reject the null hypothesis and must dig deeper to find which means are different.

To do this, I perform a tukeys test...

```{r}
#| warning: false 
#| message: false
#| echo: true
TukeyHSD(oneway)
```

## 

**Takeaway:** According to the p = .0211474 value, there is significant difference in North and West sales prices. There is no statistical evidence to support the other locations sales prices having any significant difference.

## Two Way ANOVA

**Null Hypotheses** We test two null hypothesis in this test:

1.  *There is no difference in mean Sale Price across all locations (North, South, West).*
2.  *The mean Sale Price is the same across all years houses were built in.*

**Alternative Hypotheses:**

1.  *There is a difference in mean across at least one location.*
2.  *Thrre is a difference in mean sale price accross years build.*

## 

To begin my test I grouped the year build column into decade long buckets.

```{r}
#| warning: false 
#| message: false
#| echo: true
eda2.0$YearBuilt <- paste0(floor(eda2.0$YearBuilt / 10) * 10, "-", floor(eda2.0$YearBuilt / 10) * 10 + 9)
```

## 

Next, I looked a boxplot graph that shows the means of all of our decades.

```{r}
#| warning: false 
#| message: false
#| echo: true
eda2.0 |> ggplot(aes(x = YearBuilt, y = SalePrice, fill = YearBuilt)) +
          geom_boxplot(fill = "lightblue", color = "black") +
          labs(title = "Year Built Means",
               x = "Sale Price",
               y = "Year Built") +
          theme_minimal()
```

## 

Next, I run the two way ANOVA model...

```{r}
#| warning: false 
#| message: false
#| echo: true
twoway <- aov(SalePrice ~ YearBuilt+Location, data = eda2.0)
summary(twoway)
```

## 

**Takeaway:** Both of our p values are significant! Meaning we reject both of the null hypothesis. *There is statistical evidence to support location and year build impacting sale price.*

## 

Next, I run a tukeys test to look deeper...

```{r}
#| warning: false 
#| message: false
#| echo: true
tukeys <- TukeyHSD(twoway)
print(tukeys)
```

## 

## Conclusion:

## 

**Location Takeaway:**

Houses build in the South tend to be roughly \$18,000 cheaper than houses built in the North.

Housing build in the West tend to be roughly \$13,0000 cheaper than houses built in the North.

There is no statistical evidence to support price difference from West to South houses.

## 

**Year Built Takeaways**

2010 - 2019 houses are showing the largest price variation from other decades ranging anywhere from \$150,000 - \$200,000 more than older homes.

Houses built in the 2000-2009 decade tend to be \$50,000 - \$115,000 more expensive than houses built in most time frames in the 1990s.

There is also significant evidence to support houses built in the 1990-1999 decade are priced anywhere from \$30,000 - \$100,000 higher than older homes.
